---
title: "03_Data"
format: html
---


```{r}
library(tidyverse)
```


```{r}
d <- read_csv("Data/LHDC.csv")
```

```{r}
d |> 
  filter(LS_Score != 2.3) |> 
  group_by(Population, LS_Score) |> 
  summarise(n = n())
``````

```{r}
d |> 
  filter(LS_Score != 2.3) |> 
  filter(Population == "New Zealand" & LS_Score > 1) |> 
  ggplot() + 
  aes(x=TL, colour=factor(LS_Score), group=LS_Score) +
  geom_density()
```


```{r}
d |> 
  filter(LS_Score != 2.3) |> 
  filter(Population == "South Africa" & LS_Score > 1) |> 
  ggplot() + 
  aes(x=TL, fill=factor(LS_Score), group=LS_Score) +
  geom_density(alpha = .3)
```



```{r}
d |> 
  filter(LS_Score != 2.3) |> 
  # filter(Population == "South Africa" & LS_Score > 1) |> 
  ggplot() + 
  aes(x=TL, y=after_stat(density)) +
  geom_histogram()+
  facet_grid(rows=vars(LS_Score), cols = vars(Population))
```

```{r}
d |> 
  filter(LS_Score != 2.3) |> 
  filter(Population == "South Africa" & LS_Score > 1) |> 
  ggplot() + 
  aes(x=TL) +
  geom_histogram()
```


```{r}
d |> 
  filter(Population == "New Zealand" & LS_Score > 1) |> 
  ggplot() + 
  aes(x=TL, fill=factor(LS_Score), group=LS_Score) +
  geom_histogram() +
  facet_grid(rows=vars(LS_Score))
```





