---
titlee: "Generative Classifier Models 4"
author: ""
format: 
  html:
    toc: true
---

```{r}
#| message: false
library(tidyverse)
library(rstan)
library(bayesplot)
library(posterior)
theme_set(theme_bw())
```

# Introduction

# Load functions


```{r}
source("plateau_gaussian_density.R")
```



# Set up and test model 

## Single population

### Simulate data

```{r}
# Simulate data
set.seed(12345)
n0 <- 50  # Number of samples for class 0
n1 <- 5 * n0  # Number of samples for class 1
N <- n0 + n1

# Parameters for simulation
m50_true <- 1300
sigma_x_true <- 100
d_true <- 300  # Distance factor between class 0 and 1 means

# Simulated x values
mu0_true <- m50_true - d_true / 2
mu1_true <- m50_true + d_true / 2
x0 <- c(
  rnorm(n0/2, mean = mu0_true, sd = sigma_x_true),
  rnorm(n0/2, mean = mu0_true - 250, sd = sigma_x_true*.5)
)
x1 <- c(
  rnorm(n1/2, mean = mu1_true, sd = sigma_x_true),
  rnorm(n1/2, mean = mu1_true + 250, sd = sigma_x_true*1.5)
)
x <- c(x0, x1)

# Simulated y values
y <- c(rep(0, n0), rep(1, n1))

# Data for Stan
data_list <- list(
  N = N,
  x = x,
  y = y
)
```

```{r}
data.frame(x=x,y=y) |> 
  ggplot() +
  aes(x = x, fill = factor(y), group = y) +
  geom_histogram(alpha = .5, col = NA, position = "identity") 
```

### Compile Stan Code

```{r}
stan_plateau_gaussian_twogroups <- stan_model(file = 'plateau_gaussian_twogroups.stan')
```

### Fit to simulated data

```{r}
# Fit the model using rstan
fit <- sampling(
  stan_plateau_gaussian_twogroups, 
  data = data_list,
  chains = 4,
  cores = 4,
  control = list(adapt_delta = 0.95)
)

fit
```
### Plot


```{r}
data.frame(x=x,y=y) |> 
  ggplot() +
  geom_density(
    aes(x = x, fill = factor(y), group = y),
    alpha = .5, col = NA, position = "identity"
    ) + 
  geom_line(
    data = fit |> 
      as_draws_df() |> 
      thin_draws(thin = 40) |> 
      expand_grid(x = seq(from=800, to=2100, length=500)) |> 
      rowwise() |> 
      mutate(
        p = plateau_gaussian_density(x, middle_1, delta_1, sigma_lo_1, sigma_up_1) 
      ),
    mapping = aes(x = x, y = p, group = .draw),
    alpha = .1,
    col=1
  ) + 
  geom_line(
    data = fit |> 
      as_draws_df() |> 
      thin_draws(thin = 40) |> 
      expand_grid(x = seq(from=800, to=2100, length=500)) |> 
      rowwise() |> 
      mutate(
        p = plateau_gaussian_density(x, middle_0, delta_0, sigma_lo_0, sigma_up_0) 
      ),
    mapping = aes(x = x, y = p, group = .draw),
    alpha = .1,
    col=1
  )
```


## Try on nz data

### Data

```{r}
dat <- read_csv("Data/LHDC.csv") |> 
  filter(!(Population == "Southwest Atlantic" & !is.na(OWt))) |> 
  transmute(
    Population,
    Country,
    Sex,
    Stage = as_factor(LS_Score),
    Length = TL
  )
```

### Fit model

```{r}
d_nz <- dat |> 
  filter(Length > 1050 & Length < 1650) |> 
  filter(Population == "New Zealand" & Sex == "M") |> 
  transmute(
    x = Length,
    y = as.numeric(Stage == 3 )
  )
```


```{r}
fit_d_nz <- sampling(
  stan_plateau_gaussian_twogroups, 
  data = list( N = nrow(d_nz), x = d_nz$x, y = d_nz$y ),
  chains = 4,
  cores = 4,
  control = list(adapt_delta = 0.95)
)
```

```{r}
# Extract and summarize results
print(fit_d_nz, pars = c("m50", "sigma_lo_1", "d"))
```


```{r}
# Plot posterior distribution of m50
posterior_d_nz <- extract(fit_d_nz)$m50
ggplot(data.frame(m50 = posterior_d_nz), aes(x = m50)) +
  geom_density() +
  ggtitle("Posterior distribution of m50 for NZ males") +
  theme_minimal()

```

#### Calculate probability of mature

```{r}

fit_d_nz |> 
  as_draws_df() |> 
  thin_draws(thin = 40) |> 
  expand_grid(x = seq(from=1050, to=1650, length=1000)) |> 
  rowwise() |> 
  mutate(
    d0 = plateau_gaussian_density(x, middle_0, delta_0, sigma_lo_0, sigma_up_0),
    d1 = plateau_gaussian_density(x, middle_1, delta_1, sigma_lo_1, sigma_up_1),
    n0 = d0 *
      c_plateau_gaussian_density(delta_0, sigma_lo_0, sigma_up_0),
    n1 = d1 *
      c_plateau_gaussian_density(delta_1, sigma_lo_1, sigma_up_1),
    p = n1 / ( n0 + n1 )
  ) |> 
  ggplot() + 
  geom_hline(yintercept = 0) +
  geom_line(
    mapping = aes(x = x, y = d0*100, group = .draw),
    alpha = .02,
    colour = "#00BFC4"
  ) +
  geom_line(
    mapping = aes(x = x, y = d1*100, group = .draw),
    alpha = .02,
    colour = "#F8766D"
  ) +
  geom_point(
    data = d_nz |> filter(y == 0),
    mapping = aes(x = x, y = -.08),
    shape = "|", 
    alpha = .5,
    colour = "#00BFC4"
    ) + 
  geom_point(
    data = d_nz |> filter(y == 1),
    mapping = aes(x = x, y = -.04),
    shape = "|", 
    alpha = .5,
    colour = "#F8766D"
    ) + 
  geom_line(
    mapping = aes(x = x, y = p, group = .draw),
    alpha = .2
    ) +
  # xlim(1050,1550) + 
  ylab("Posterior probability of maturity") + 
  ggtitle("NZ data")

```



### Try on nz data with cutpoints 900 and 1650

#### Fit model

```{r}
d_nz2 <- dat |> 
  filter(Length > 900 & Length < 1650) |> 
  filter(Population == "New Zealand" & Sex == "M") |> 
  transmute(
    x = Length,
    y = as.numeric(Stage == 3 )
  )
```


```{r}
fit_d_nz2 <- sampling(
  stan_plateau_gaussian_twogroups, 
  data = list( N = nrow(d_nz2), x = d_nz2$x, y = d_nz2$y ),
  chains = 4,
  cores = 4,
  control = list(adapt_delta = 0.95)
)
```

```{r}
# Extract and summarize results
print(fit_d_nz2)
```


```{r}
# Plot posterior distribution of m50
posterior_d_nz2 <- extract(fit_d_nz2)$m50
ggplot(data.frame(m50 = posterior_d_nz2), aes(x = m50)) +
  geom_density() +
  ggtitle("Posterior distribution of m50 for NZ males") +
  theme_minimal()

```

#### Calculate probability of mature

```{r}

fit_d_nz2  |> 
  as_draws_df() |> 
  thin_draws(thin = 40) |> 
  expand_grid(x = seq(from=900, to=1650, length=1000)) |> 
  rowwise() |> 
  mutate(
    d0 = plateau_gaussian_density(x, middle_0, delta_0, sigma_lo_0, sigma_up_0),
    d1 = plateau_gaussian_density(x, middle_1, delta_1, sigma_lo_1, sigma_up_1),
    n0 = d0 *
      c_plateau_gaussian_density(delta_0, sigma_lo_0, sigma_up_0),
    n1 = d1 *
      c_plateau_gaussian_density(delta_1, sigma_lo_1, sigma_up_1),
    p = n1 / ( n0 + n1 )
  ) |> 
  ggplot() + 
  geom_hline(yintercept = 0) +
  geom_line(
    mapping = aes(x = x, y = d0*100, group = .draw),
    alpha = .02,
    colour = "#00BFC4"
  ) +
  geom_line(
    mapping = aes(x = x, y = d1*100, group = .draw),
    alpha = .02,
    colour = "#F8766D"
  ) +
  geom_point(
    data = d_nz2 |> filter(y == 0),
    mapping = aes(x = x, y = -.08),
    shape = "|", 
    alpha = .5,
    colour = "#00BFC4"
    ) + 
  geom_point(
    data = d_nz2 |> filter(y == 1),
    mapping = aes(x = x, y = -.04),
    shape = "|", 
    alpha = .5,
    colour = "#F8766D"
    ) + 
  geom_line(
    mapping = aes(x = x, y = p, group = .draw),
    alpha = .2
    ) +
  # xlim(1050,1550) + 
  ylab("Posterior probability of maturity") + 
  ggtitle("NZ data (truncated 900 to 1650)")



```


```{r}
ggplot() + 
  geom_hline(yintercept = 0) +
  geom_density(
    data = d_nz2,
    mapping = aes(x = x, colour = factor(y), group = factor(y)),
  ) +
  geom_point(
    data = d_nz2 |> filter(y == 0),
    mapping = aes(x = x, y = -.0008),
    shape = "|", 
    alpha = .5,
    colour = "#00BFC4"
    ) + 
  geom_point(
    data = d_nz2 |> filter(y == 1),
    mapping = aes(x = x, y = -.0004),
    shape = "|", 
    alpha = .5,
    colour = "#F8766D"
    ) + 
  # xlim(1050,1550) + 
  ylab("Posterior probability of maturity") + 
  ggtitle("NZ data (truncated 900 to 1650)")
```


### Try on nz data with cutpoints 400 and 1650

#### Fit model

```{r}
d_nz3 <- dat |> 
  filter(Length > 400 & Length < 1650) |> 
  filter(Population == "New Zealand" & Sex == "M") |> 
  transmute(
    x = Length,
    y = as.numeric(Stage == 3 )
  )
```


```{r}
fit_d_nz3 <- sampling(
  stan_plateau_gaussian_twogroups, 
  data = list( N = nrow(d_nz3), x = d_nz3$x, y = d_nz3$y ),
  chains = 4,
  cores = 4,
  control = list(adapt_delta = 0.95)
)
```

```{r}
# Extract and summarize results
print(fit_d_nz3)
```

## J populations

### Compile Stan code

```{r}
stan_gen04_populations <- stan_model(file = 'gen04_populations.stan')
```
### Simulate data

```{r}

# Simulate some data
set.seed(123)
N <- 1000  # Total number of data points
J <- 5    # Number of populations
pop <- sample(1:J, N, replace = TRUE)

# Generate data for each population

mu_m50 <- 1300
d <- 200
mu_mu0 <- mu_m50 - d / 2
mu_mu1 <- mu_m50 + d / 2
sigma_alpha <- 20
sigma_x <- 50

mu0_pop <- rnorm(J, mu_mu0, sigma_alpha)
mu1_pop <- rnorm(J, mu_mu1, sigma_alpha)

x <- numeric(N)
y <- rbinom(N, 1, 0.5)

for (i in 1:N) {
  if (y[i] == 0) {
    x[i] <- rnorm(1, mean=mu0_pop[pop[i]], sd=sigma_x)
  } else {
    x[i] <- rnorm(1, mean=mu1_pop[pop[i]], sd=sigma_x)
  }
}


# Prepare data for Stan
data_list_populations <- list(
  N = N,
  x = x,
  y = y,
  J = J,
  pop = pop
)
```


### Fit to simulated data

```{r}
# Fit the model using rstan
fit_populations <- sampling(
  stan_gen04_populations, 
  data = data_list_populations,
  chains = 4,
  # iter = 2000,
  # warmup = 500,
  cores = 4,
  control = list(adapt_delta = 0.95)
)
```



# Maturity

## Fit to actual data

### Prepare data 

```{r}
d_pop <- dat |> 
  filter(Length > 1050 & Length < 1550 & Sex == "M" & Population != "Northeast Pacific") |> 
  transmute(
    x = Length,
    y = as.numeric(Stage == 3),
    pop = Population |> as.factor() |> as.numeric()
  )
```

```{r}
popkey <- dat |> 
  filter( Population != "Northeast Pacific" ) |> 
  transmute(
    pop = Population |> as.factor() |> as.numeric(),
    key = Population
  ) |> 
  unique() |> 
  arrange(pop)

popkey
```

### Fit model

```{r}
fit_d_pop <- sampling(
  stan_gen04_populations, 
  data = list( 
    N = nrow(d_pop), 
    x = d_pop$x, 
    y = d_pop$y, 
    pop = d_pop$pop, 
    J = n_distinct(d_pop$pop),
    
    prior_mu_m50_mu = 1300,
    prior_mu_m50_sigma = 100,
  
    prior_d_mu = 200,
    prior_d_sigma = 50,
  
    prior_sigma_x_sigma = 50,
    prior_sigma_alpha_sigma = 50
    
    ),
  chains = 4,
  # iter = 2000,
  # warmup = 500,
  cores = 4,
  control = list(adapt_delta = 0.95)
)
```


```{r}
fit_d_pop
```

```{r}
popkey |>
  bind_cols(m50 = c(1292.91,1289.70,1274.39,1285.80,1238.32 ))
```

### Checking

```{r}
check_hmc_diagnostics(fit_d_pop)
```
```{r}
fit_d_pop |> 
  mcmc_pairs(
    pars = c(
      "mu_m50", 
      # "m50_pop[1]", "m50_pop[2]","m50_pop[3]","m50_pop[4]","m50_pop[5]","m50_pop[6]",
      "d", "sigma_x", "sigma_alpha")
    )
```

```{r}
#| fig-width: 20
#| fig-height: 20

fit_d_pop |> 
  mcmc_pairs(
    pars = c(
      "mu_m50",
      "m50_pop[1]", "m50_pop[2]","m50_pop[3]","m50_pop[4]","m50_pop[5]"
      # "d", "sigma_x", "sigma_alpha"
      )
    )
```

```{r}
#| fig-width: 20
#| fig-height: 20

fit_d_pop |> 
  mcmc_pairs(
    pars = c(
      "mu_m50",
      "alpha[1]", "alpha[2]","alpha[3]","alpha[4]","alpha[5]",
      "d",
      "sigma_x",
      "sigma_alpha"
      )
    )
```



```{r}
mcmc_areas(
  fit_d_pop |> as.matrix(),
  pars = c("mu_m50", "m50_pop[1]", "m50_pop[2]", "m50_pop[3]", "m50_pop[4]", "m50_pop[5]"),
  prob = 0.8
  ) +
  scale_y_discrete(
    labels = c(
      "mu_m50" = "Overall mean",
      "m50_pop[1]" = popkey$key[1], 
      "m50_pop[2]" = popkey$key[2], 
      "m50_pop[3]" = popkey$key[3], 
      "m50_pop[4]" = popkey$key[4], 
      "m50_pop[5]" = popkey$key[5])) +
  xlab("Transition length")
  
```

# Birth


```{r}
#| fig-height: 8
#| fig-width: 6

dat |> 
  filter( Length < 500 & !is.na(Sex) ) |>
  ggplot() +
  aes(x = Length, fill = factor(Stage), group = Stage) +
  geom_density(alpha = .5, col = NA) + 
  facet_grid(
    cols = vars(Sex),
    rows = vars(Population),
    scales = "free_y"
    ) 
```


```{r}
#| fig-height: 8
#| fig-width: 6

dat |> 
  filter( Length < 500 & !is.na(Sex) ) |>
  ggplot() +
  aes(x = Length, fill = factor(Stage), group = Stage) +
  geom_histogram(alpha = .5, col = NA, position = "identity") + 
  facet_grid(
    cols = vars(Sex),
    rows = vars(Population)
    ) 
```


```{r}
#| fig-height: 8
#| fig-width: 6

dat |> 
  filter( Length > 150 & Length < 450 ) |>
  ggplot() +
  aes(x = Length, fill = factor(Stage), group = Stage) +
  geom_histogram(alpha = .5, col = NA, position = "identity") + 
  facet_grid(
    rows = vars(Population)
    ) 
```
